A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol

### 2024-01-01 paper.tex
reduce 2 pages

Our experiment was conducted over three data collection periods to
investigate the difference between conventional hypothesis testing and
visual inference in the application of linear regression diagnostics.
Two types of departures, non-linearity and heteroskedasticity, were
collected during data collection periods I and II. The data collection
period III was designed primarily to measure human responses to null
lineups so that the visual \(p\)-values can be estimated. Additional
lineups for both non-linearity and heteroskedasticity, using uniform
fitted value distributions, were included for additional data, and to
avoid participant frustration of too many difficult tasks. Overall, we
collected 7974 evaluations on 1152 unique lineups performed by 443
participants throughout three data collection periods.

@@

\hypertarget{non-linearity}{%
\subsubsection{Non-linearity}\label{non-linearity}}

@@

@@

\begin{align*} \label{eq:nonlinearity-model}
\boldsymbol{y} &= \boldsymbol{1}_n + \boldsymbol{x} + \boldsymbol{z} + \boldsymbol{\varepsilon},\\
\boldsymbol{x} &= g(\boldsymbol{x}_{raw}, 1), \\
\boldsymbol{z} &= g(\boldsymbol{z}_{raw}, 1), \\
\boldsymbol{z}_{raw} &= He_j(g(\boldsymbol{x}, 2)),
\end{align*}

\noindent where \(\boldsymbol{y}\), \(\boldsymbol{x}\),
\(\boldsymbol{\varepsilon}\), \(\boldsymbol{x}_{raw}\),
\(\boldsymbol{z}_{raw}\) are vectors of size \(n\), \(\boldsymbol{1}_n\)
is a vector of ones of size \(n\), \(He_{j}(.)\) is the \(j\)th-order
probabilist's Hermite polynomials
\citetext{\citealp{hermite1864nouveau}; \citealp[originally
by][]{de1820theorie}},
\(\varepsilon \sim N(\boldsymbol{0}_n, \sigma^2\boldsymbol{I}_n)\), and
\(g(\boldsymbol{x}, k)\) is a scaling function to enforce the support of
the random vector to be \([-k, k]^n\) defined as

\begin{equation} \label{eq:scaling-function}
g(\boldsymbol{x}, k) = 2k\cdot\frac{\boldsymbol{x} - x_{min}\boldsymbol{1}_n}{x_{max} - x_{min}} - k, \quad \text{for} \quad k > 0. 
\end{equation}

\noindent where
\(\displaystyle x_{min} = \min_{i\in \{1, ..., n\}} x_i\),
\(\displaystyle x_{max} = \max_{i\in \{1, ..., n\}} x_i\) and \(x_i\) is
the \(i\)-th entry of \(\boldsymbol{x}\). The function \texttt{hermite}
from the R package \texttt{mpoly} \citep{mpoly} is used to simulate
\(\boldsymbol{z}_{raw}\) to generate Hermite polynomials.

The null regression model used to fit the realizations generated by the
above model is formulated as

\begin{equation} \label{eq:null-model}
\boldsymbol{y} = \beta_0\boldsymbol{1}_n + \beta_1 \boldsymbol{x} + \boldsymbol{u},
\end{equation}

\noindent where
\(\boldsymbol{u} \sim N(\boldsymbol{0}, \sigma^2\boldsymbol{I}_n)\).
Here \(\boldsymbol{z}\) is a higher order term of \(\boldsymbol{x}\) and
leaving it out in the null regression results in model misspecification.

@@

Visual patterns of non-linearity are simulated using four different
orders of probabilist's Hermite polynomials (\(j = 2, 3, 6, 18\)). The
values of \(j\) were chosen so that distinct shapes of non-linearity
were included in the residual plot. These include ``U'', ``S'', ``M''
and ``triple-U'' shape as shown in Figure
\ref{fig:different-shape-of-herimite}. A greater value of \(j\) will
result in a curve with more turning points. It is expected that the
``U'' shape will be the easiest to detect, and as the shape gets more
complex it will be harder to perceive in a scatterplot, particularly
when there is noise. Figure \ref{fig:different-sigma} shows the ``U''
shape for different amounts of noise (\(\sigma\)).

\hypertarget{heteroskedasticity}{%
\subsubsection{Heteroskedasticity}\label{heteroskedasticity}}

Data collection period II was designed to study the ability of
participants to detect the heteroskedasticity under a simple linear
regression model setting:

\begin{align*} \label{eq:heter-model}
\boldsymbol{y} &= \boldsymbol{1}_n + \boldsymbol{x} + \boldsymbol{\varepsilon},\\
\boldsymbol{x} &= g(\boldsymbol{x}_{raw}, 1),\\
\boldsymbol{\varepsilon} &\sim N(\boldsymbol{0}_n, \boldsymbol{1}_n + (2 - |a|)(\boldsymbol{x} - a\boldsymbol{1}_n)'(\boldsymbol{x} - a\boldsymbol{1}_n) b \boldsymbol{I}_n), 
\end{align*}

\noindent where \(\boldsymbol{y}\), \(\boldsymbol{x}\),
\(\boldsymbol{\varepsilon}\) are vectors of size \(n\) and \(g(.)\) is
the scaling function defined in Equation \ref{eq:scaling-function}. The
null regression model used to fit the realizations generated by the
above model is formulated exactly the same as Equation
\ref{eq:null-model}.

For \(b \neq 0\), the variance-covariance matrix of the error term
\(\boldsymbol{\varepsilon}\) is correlated with the predictor
\(\boldsymbol{x}\), which will lead to the presence of
heteroskedasticity. Visual patterns of heteroskedasticity are simulated
using three different shapes (\(a\) = -1, 0, 1).

Since \(\text{supp}(X) = [-1, 1]\), choosing \(a\) to be \(-1\), \(0\)
and \(1\) can generate ``left-triangle'', ``butterfly'' and
``right-triangle'' shapes as displayed in Figure
\ref{fig:different-shape-of-heter}. The term \((2 - |a|)\) maintains the
magnitude of residuals across different values of \(a\). Figure
\ref{fig:different-b} shows the butterfly shape as the ratio parameter
(\(b\)) is changed.

@@

\caption{Five different values of $b$ are used in heteroskedasticity simulation to control the strength of the signal. Larger values of $b$ yield a bigger difference in variation, and stus stronger heteroskedasticity signal.}\label{fig:different-b}
\end{figure}

@@

Fitted values are a function of the independent variables, and the
distribution of the observed values affects the distribution of the
fitted values. In the best case scenario the fitted values will have a
uniform distribution, which means that there is even coverage of
possible observed values across all of the predictors. This is not
always present in the collected data. Sometimes the fitted values are
discrete because one or more predictors were measured discretely. It is
also common to see a skewed distribution of fitted values if one or more
of the predictors has a skewed distribution. This latter problem is
usually corrected before modelling, using a variable transformation. Our
simulation assess this by using four different distributions to
represent fitted values, constructed by different sampling of the raw
predictor \(X_{raw}\):

\begin{itemize}
\tightlist
\item
  uniform, \(U(-1, 1)\),
\item
  normal, \(N(0, 0.3^2)\),
\item
  skewed, \(\text{lognormal}(0, 0.6^2)/3\), and
\item
  discrete, \(U\{1, 5\}\).
\end{itemize}

\noindent Figure \ref{fig:different-dist} shows the non-linear pattern,
a ``U'' shape, with the different fitted value distributions. We would
expect that structure in residual plots would be easier to perceive when
the fitted values are uniformly distributed.

@@

\hypertarget{controlling-the-signal-strength}{%
\subsubsection{Controlling the signal
strength}\label{controlling-the-signal-strength}}

The three parameters \(n\), \(\sigma\) and \(b\) are important for
controlling the strength of the signal to generate lineups with a
variety difficulty levels. This will ensure that estimated power curves
will be smooth and continuous, and that participants are allocated a set
of lineups with a range of difficulty.

Parameter \(\sigma \in \{0.5, 1, 2, 4\}\) and
\(b \in \{0.25, 1, 4, 16, 64\}\) are used in data collection periods I
and II respectively. A large value of \(\sigma\) will increase the
variation of the error of the non-linearity model and decrease the
visibility of the visual pattern. The parameter \(b\) controls the
variation in the standard deviation of the error across the support of
the predictor. A larger value of \(b\) generates a larger ratio between
it's smallest and highest values, making the visual pattern more
obvious. The sample size \(n\) sharpens (or blurs) a pattern when it is
larger (or smaller). Figures \ref{fig:different-sigma},
\ref{fig:different-b} and \ref{fig:different-n} demonstrate the impact
of these parameters on signal strength.

\hypertarget{experimental-setup}{%
\subsection{Experimental setup}\label{experimental-setup}}

@@

\hypertarget{effect-size}{%
\subsubsection{Effect size}\label{effect-size}}

@@

For the non-linearity model, the key items defining effect size are
sample size (\(n\)) and variance of the error term (\(\sigma^2\)), and
so effect size would be roughly calculated as \(\sqrt{n}/{\sigma}\). As
sample size increases the effect size would increase, but as variance
increases the effect size decreases. However, it is not clear how the
additional parameter for the model polynomial order, \(k\), should be
incorporated. Intuitively, the large \(k\) means more complex pattern,
which likely means effect size would decrease. For the purposes of our
calculations we have chosen to use an approach based on Kullback-Leibler
divergence \citep{kullback1951information}, coupled with simulation.
This formulation defines effect size to be:

@@

\hypertarget{lineup-allocation-to-participants}{%
\subsubsection{Lineup allocation to
participants}\label{lineup-allocation-to-participants}}

From Table \ref{tab:model-factor-table} it can be computed that there
are a total of \(4 \times 4 \times 3 \times 4 = 192\) and
\(3 \times 5 \times 3 \times 4 = 180\) combinations of parameter values
for the non-linearity model and heteroskedasticity models respectively.
Three replications for each combination results in
\(192 \times 3 = 576\) and \(180 \times 3 = 540\) lineups, respectively.

Each lineup needs to be evaluated by at least five participants. From
previous work, and additional pilot studies for this experiment, we
decided to record evaluations from 20 lineups for each participant. Two
of the 20 lineups with clear visual patterns were used as attention
checks, to ensure quality data. Thus, \(576 \times 5 / (20-2) = 160\)
and \(540 \times 5 / (20-2) = 150\) participants were needed to cover
the experimental design for the data collection periods I and II,
respectively. The factor levels and range of difficulty was assigned
relatively equally among participants. The Appendix contains graphical
summaries of these assignments for each subject.

Data collection period III was primarily to obtain Rorschach lineup
evaluations to estimate \(\alpha\) (see Section A.1 in the Appendix),
and also to obtain additional evaluations of lineups made with uniform
fitted value distribution to ensure consistency in the results. To
construct a Rorschach lineup, the data is generated from a model with
zero effect size, while the null data are generated using the same
simulation method discussed in Section
\ref{visual-test-procedure-based-on-lineups}. This procedure differs
from that of the canonical Rorschach lineup, where all 20 plots are
generated directly from the null model, so the method suggested in
\citet{vanderplas2021statistical} for typical lineups containing a data
plot is used to estimate \(\alpha\). (The Appendix contains a
sensitivity analysis on the effect of uncertainty in the estimation of
\(\alpha\).) The \(3 \times 4 = 12\) treatment levels of the common
factors, replicated three times results in 36 lineups. And 6 more
evaluations on the 279 lineups with uniform fitted value distribution,
results in at least
\((36 \times 20 + 279 \times 3 \times 6) / (20-2) = 133\) participants
needed.

\hypertarget{collecting-results}{%
\subsubsection{Collecting results}\label{collecting-results}}

Participants for all three data collection periods were recruited from
the Prolific crowd-sourcing platform \citep{palan2018prolific}.
Pre-screening procedures were applied during the recruitment:
participants were required to be fluent in English, with \(98\%\)
minimum approval rate and at least 10 submissions in other studies.

During the experiment, every participant is presented with a block of 20
lineups. A lineup consists of a randomly placed data plot and 19 null
plots, which are all residual plots drawn with raw residuals on the
y-axis and fitted values on the x-axis. An additional horizontal red
line is added at \(y = 0\) as a visual reference. The data in the data
plot is simulated from one of two models described in Section
\ref{simulating-departures-from-good-residuals}, while the data of the
remaining 19 null plots are generated by the residual rotation technique
discussed in Section \ref{visual-test-procedure-based-on-lineups}.

>>>>

Our experiment was conducted over three data collection periods to
investigate the difference between conventional hypothesis testing and
visual inference in the application of linear regression diagnostics.
Two types of departures, non-linearity and heteroskedasticity, were
collected during data collection periods I and II. The data collection
period III was designed primarily to measure human responses to null
lineups so that the visual \(p\)-values can be estimated. Additional
lineups for both non-linearity and heteroskedasticity, using uniform
fitted value distributions, were included for additional data, and to
avoid participant frustration of too many difficult tasks.

During the experiment, every participant recruited from the Prolific
crowd-sourcing platform \citep{palan2018prolific} is presented with a
block of 20 lineups. A lineup consists of a randomly placed data plot
and 19 null plots, which are all residual plots drawn with raw residuals
on the y-axis and fitted values on the x-axis. An additional horizontal
red line is added at \(y = 0\) as a visual reference. The data in the
data plot is simulated from one of two models described in Section
\ref{simulating-departures-from-good-residuals}, while the data of the
remaining 19 null plots are generated by the residual rotation technique
discussed in Section \ref{visual-test-procedure-based-on-lineups}.

Overall, we collected 7974 evaluations on 1152 unique lineups performed
by 443 participants throughout three data collection periods. Details
about lineup allocation and data collection are provided in the
Appendix.

@@

@@

\hypertarget{non-linearity-and-heteroskedasticity}{%
\subsubsection{Non-linearity and
Heteroskedasticity}\label{non-linearity-and-heteroskedasticity}}

@@

\[ \boldsymbol{y} = \boldsymbol{1}_n + \boldsymbol{x} + \boldsymbol{z} + \boldsymbol{\varepsilon},~\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \sigma^2\boldsymbol{I}_n)\]

\noindent where \(\boldsymbol{y}\), \(\boldsymbol{x}\),
\(\boldsymbol{\varepsilon}\), \(\boldsymbol{z}\) are vectors of size
\(n\), \(\boldsymbol{1}_n\) is a vector of ones of size \(n\),
\(\boldsymbol{z} \propto He_j(\boldsymbol{x})\), \(He_{j}(.)\) is the
\(j\)th-order probabilist's Hermite polynomials
\citetext{\citealp{hermite1864nouveau}; \citealp[originally
by][]{de1820theorie}}.

To fit a misspecified model, the \(\boldsymbol{z}\) term, which is a
higher order term of \(\boldsymbol{x}\), was omitted from the null
regression formula

\begin{equation} \label{eq:null-model}
\boldsymbol{y} = \beta_0\boldsymbol{1}_n + \beta_1 \boldsymbol{x} + \boldsymbol{u}, \boldsymbol{u} \sim N(\boldsymbol{0}_n, \sigma^2\boldsymbol{I}_n).
\end{equation}

Four different values of \(j = 2, 3, 6, 18\) were chosen so that
distinct shapes of non-linearity were included in the residual plot.
These include ``U'', ``S'', ``M'' and ``triple-U'' shape as shown in
Figure \ref{fig:different-shape-of-herimite}. A greater value of \(j\)
will result in a curve with more turning points. It is expected that the
``U'' shape will be the easiest to detect, and as the shape gets more
complex it will be harder to perceive in a scatterplot, particularly
when there is noise. Figure \ref{fig:different-sigma} shows the ``U''
shape for different amounts of noise (\(\sigma\)).

Data collection period II was similar to Data collection I but focuses
on heteroskedasticity visual patterns formulated as

\[\boldsymbol{y} = \boldsymbol{1}_n + \boldsymbol{x} + \boldsymbol{\varepsilon},~\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \boldsymbol{1}_n + (2 - |a|)(\boldsymbol{x} - a\boldsymbol{1}_n)'(\boldsymbol{x} - a\boldsymbol{1}_n) b \boldsymbol{I}_n),\]
\noindent where \(\boldsymbol{y}\), \(\boldsymbol{x}\),
\(\boldsymbol{\varepsilon}\) are vectors of size \(n\).

The null regression model used to fit the realizations generated by the
above model is formulated exactly the same as Equation
\ref{eq:null-model}. For \(b \neq 0\), the variance-covariance matrix of
the error term \(\boldsymbol{\varepsilon}\) is correlated with the
predictor \(\boldsymbol{x}\), which will lead to the presence of
heteroskedasticity.

Visual patterns of heteroskedasticity are simulated using three
different shapes (\(a\) = -1, 0, 1) including ``left-triangle'',
``butterfly'' and ``right-triangle'' shapes as displayed in Figure
\ref{fig:different-shape-of-heter}. Figure \ref{fig:different-b} shows
the butterfly shape as the ratio parameter (\(b\)) is changed.

More details about the design of the data generating process are
provided in the Appendix.

@@

@@

\caption{Five different values of $b$ are used in heteroskedasticity simulation to control the strength of the signal. Larger values of $b$ yield a bigger difference in variation, and thus stronger heteroskedasticity signal.}\label{fig:different-b}
\end{figure}

@@

Fitted values are a function of the independent variables, and the
distribution of the observed values affects the distribution of the
fitted values. In the best case scenario the fitted values will have a
uniform distribution, which means that there is even coverage of
possible observed values across all of the predictors. This is not
always present in the collected data. Sometimes the fitted values are
discrete because one or more predictors were measured discretely. It is
also common to see a skewed distribution of fitted values if one or more
of the predictors has a skewed distribution. This latter problem is
usually corrected before modelling, using a variable transformation. Our
simulation assess this by using four different distributions to
represent fitted values, constructed by different sampling of the raw
predictor \(X_{raw}\), namely, uniform (\(U(-1, 1)\)), normal
(\(N(0, 0.3^2)\)), skewed (\(\text{lognormal}(0, 0.6^2)/3\)) and
discrete (\(U\{1, 5\}\)).

Figure \ref{fig:different-dist} shows the non-linear pattern, a ``U''
shape, with the different fitted value distributions. We would expect
that structure in residual plots would be easier to perceive when the
fitted values are uniformly distributed.

@@

\hypertarget{effect-size}{%
\subsection{Effect size}\label{effect-size}}

@@

@@

For the non-linearity model, the key items defining effect size are
sample size (\(n\)) and variance of the error term (\(\sigma^2\)), and
so effect size would be roughly calculated as \(\sqrt{n}/{\sigma}\). As
sample size increases the effect size would increase, but as variance
increases the effect size decreases. However, it is not clear how the
additional parameter for the model polynomial order, \(j\), should be
incorporated. Intuitively, the large \(j\) means more complex pattern,
which likely means effect size would decrease. For the purposes of our
calculations we have chosen to use an approach based on Kullback-Leibler
divergence \citep{kullback1951information}, coupled with simulation.
This formulation defines effect size to be:

@@
